---
title: "Practice: Introduction to ggplot2"
subtitle: "Data Visualization with R and ggplot2"
author: "Johan Larsson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Overview

For this session, you will need the following packages:

- [gganimate](https://cran.r-project.org/package=gganimate)
- [plotly](https://cran.r-project.org/package=plotly)
- [gapminder](https://cran.r-project.org/package=gapminder)
- [transformr](https://cran.r-project.org/package=transformr)

Install them by calling

```{r install-packages, eval = FALSE}
install.packages(c(
  "gapminder",
  "gganimate",
  "plotly",
  "transformr"
))
```

## Renderer

For **gganimate** to be able to produce animated images, you need to install a
renderer. [gifski](https://gif.ski/) is usually the best choice, but can be
somewhat tricky to install depending on what operating system you are on. If you
aren't able to install gifski for some reason, you can also use
[ImageMagick](https://imagemagick.org).

First, simply try to run 

```{r install-gifski, eval = FALSE}
install.packages("gifski")
```

or 

```{r install-magick, eval = FALSE}
install.packages("magick")
```

If any of the above works without a hitch, then you're done! If not, try the 
advanced instructions below.

### Advanced Instructions for Windows

Try to install **imagemagick** by downloading and installing the recommended
installation file at <https://imagemagick.org/script/download.php#windows>.

After this try to run `install.packages("magick")` again.

### Advanced Instructions for macOS

To install **gifski** on mac, first make sure you make **homebrew** installed.
If you do not, open the macOS terminal and run 

```bash
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
```

to install it. Afterwards, you call

```bash
brew install rustc
```

and then install the **gifski** R package by running

```r
install.packages("gifski")
```

as before.

You can also try to install the Mac App Store application
[here](https://sindresorhus.com/gifski), if the solution above does not work. 

To install **ImageMagick** on macOS, you can do so using **homebrew**. First
make sure you've installed **homebrew** (as outlined above) and then run

```bash
brew install imagemagick
```

### Linux

Install **gifski** is easy using snap. Simply go to <https://snapcraft.io/gifski>
for instructions for any given distribution.

# Gapminder

Do you remember the animated plots we produced in the introductory
presentation workshop based on the Gapminder Hans Rosling
animated visualization?

```{r rosling, echo = FALSE, out.extra='style="border: none;"', out.width="100%"}
knitr::include_url("https://www.youtube.com/embed/jbkSRLYSojo")
```

In this worked example, we'll work out how to
reproduce that plot as both an animated an interactive visualization.

# Animated Visualization

The dataset that we'll use is available via the
[gapminder](https://CRAN.R-project.org/package=gapminder) package. Loadign
**gapminder** mThisakes the dataset directly available in an object called
`gapminder`. These are the first few rows of the dataset.

```{r}
library(gapminder)

head(gapminder)
```

The variables should be self-explanatory.

Let's jump right in and create a bubble plot faceted on year (which we cut
into groups), with population mapped to the size of the bubbles, and
GDP per capita and life expectancy on the x and y axes respectively.

```{r fig.cap = "Life Expectancy and GDP per capita from 1950 to 2010.", fig.width = 8, fig.height = 7}
library(tidyverse)

gapminder %>%
  mutate(years = cut_interval(year, length = 5)) %>%
  ggplot(aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap("years") +
  labs(
    y = "Life Expectancy",
    x = "GDP per Capita",
    size = "population"
  )
```

Just as I said in the first presentation, this visualization is not (yet) working
out so well for us. Let's make it animated instead. For this, we'll use
the [gganimate](https://CRAN.R-project.org/package=gganimate) package.
First install the package.

```{r, eval = FALSE}
install.packages("gganimate")
```

To use the **gganimate** package you also need a *renderer* to produce
animated images. You can use either [gifski](https://gif.ski/) or
[ImageMagick](https://imagemagick.org/). We recommend the former (and
**gganimate** defaults to gifski if it is installed), but either
will work just fine. Run one (or both) of the following lines to
install a renderer.

```{r, eval = FALSE}
install.packages("gifski")
install.packages("magick")
```

We build the plot as before, but now make it animated by adding
the `transition_time()` function to the plot as well as use the
title label to show the current year.

```{r, fig.cap = "GDP per capita and life expectancy for some of the countries of the world.", fig.width = 7, fig.height = 6}
library(gganimate)

ggplot(gapminder,  aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.5) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  labs(
    title = "Year: {frame_time}", # special glue syntax,
    y = "Life Expectancy",
    x = "GDP per Capita",
    size = "population"
  ) +
  transition_time(year)
```

If you think the plot is still crowded, we could alternatively use
facets to separate continents. Here we also make use of the
`country_colors` object that is included in the **gapminder** package.

```{r, fig.cap = "GDP per capita and life expectancy; now with facets!", fig.width = 7, fig.height = 6}
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(
    title = "Year: {frame_time}",
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  transition_time(year)
```

So far our plot does a good job of showing the trends among the various
continents of the world but is hard to use if we are interested in one
specific country. A remedy for this can be to use labels to let us
identify which bubble belongs to which country. The large number of countries,
however, means that it's not a frightfully good idea to label *all* of them.

Instead, we'll pick out the largest two countries (at the latest
time stamp) on each continent and label those. First, we
store the names of the countries in a vector, `large_country_names`.

The following steps first filter the dataset so that only
observations from the latest year (`max(year)`) are kept, then groups the
dataset by continent, then slices the dataset so that the observations
(countries) with the largest and next-to-largest values of population (`pop`)
of each group (continent) are kept, and then finally pulls out (using `pull()`)
the country names.

```{r}
large_country_names <-
  gapminder %>%
  filter(year == max(year)) %>%
  group_by(continent) %>%
  slice_max(pop, n = 2) %>%
  pull(country)

large_country_names
```

Then we filter the original dataset to create a separate dataset for our
labels.

```{r}
large_countries <- filter(gapminder, country %in% large_country_names)
```

Now we put everything together; this time we also change the
*easing* of aesthetics from linear to cubic in-and-out using `ease_aes()`,
to more clearly show that we actually only have data on a 5-year interval here.
We label the countries with `geom_label_repel()` from the **ggrepel** package,
in order to avoid overlapping labels.^[Working with labels and animated
visualizations is something of a challenge. Here I've had to tweak the
settings (mostly `nudge_x` and `nudge_y`) several times in order to get
something that looks good.]

```{r, fig.cap = "Life expectancy and GDP per capita with countries. The two largest countries at the start (in terms of population) have been labeled.", fig.width = 8, fig.height = 7}
library(ggrepel)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(alpha = 0.5) +
  geom_label_repel(
    aes(gdpPercap, lifeExp, label = country),
    inherit.aes = FALSE,
    seed = 1, # important when animating
    nudge_x = 5,
    nudge_y = -10,
    data = large_countries
  ) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(
    title = "Year: {frame_time}",
    x = "GDP per capita",
    y = "Life expectancy"
  ) +
  transition_time(year) +
  ease_aes("cubic-in-out")
```

# Interactive Visualization

Interactive visualizations are often effective, particularly when
we want to visualize a complicated dataset such as this one.
Here we'll use the [plotly](https://CRAN.R-project.org/package=plotly) package
to do so, which, as you may recall from the lecture, works well in tandem
with ggplot. First install the package.

```{r, eval = FALSE}
install.packages("plotly")
```

Then load the package.

```{r}
library(plotly)
```

Now we redraw the plot, adding an interactive slider to select the year using
plotly. Make note of the additional mapping that we've
added to `geom_point()`, namely `frame`, which is a special
mapping that will let plotly know which variable to use to separate
the visualization into frames.

```{r, fig.cap = "An interactive visualization using plotly for the Gapminder data.", fig.width = 8, fig.height = 7}
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, color = country)) +
  geom_point(aes(frame = year), alpha = 0.5) +
  scale_colour_manual(values = country_colors, guide = FALSE) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(x = "GDP per capita", y = "Life expectancy")

ggplotly(p)
```

Notice how seamless the conversion of ggplots into interactive plots can be
with the help of plotly.
