---
title: "Practice: Categorical Data and Color"
subtitle: "Data Visualization with R and ggplot2"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = FALSE,
  fig.width  = 5,
  fig.height = 4,
  dev        = "png"
)

options(scipen = 999)

library(tidyverse)
```

# Overview

For this quiz you are going to need the
[titanic](https://CRAN.R-project.org/package=titanic) package

If you haven't already installed it, just call

```{r eval = FALSE}
install.packages("titanic")
```

# The Maiden Voyage of the Titanic

Let's explore some date of survival from the maiden voyage of the Titanic. There
is a data set available in the base R distribution, but we'll use a slightly
more extensive data set from the **titanic** package. This data set is divided
into two objects: `titanic_train` and `titanic_test`. The latter of these,
however, does not contain information on survival, so we will just use the
`titanic_train` set.

```{r}
library(titanic)
```

Call `head()` on the data set (`titanic_train`) to get an idea of what it looks
like.

Well, it's not very neatly formatted. Tidy it up by calling the following code
in which we rename and re-code some of the variables. 

```{r}
titanic <-
  as_tibble(titanic_train) |>
  select(PassengerId, Class = Pclass, Sex, Age, Survived) |>
  mutate(
    Age = as_factor(ifelse(Age < 18, "Chid", "Adult")),
    Survived = as_factor(ifelse(Survived, "Yes", "No"))
  )
```

## A Simple Bar Chart

Let's start easy: begin with a simple bar chart of the counts of each class.

```{r include = FALSE}
ggplot(titanic, aes(Class)) +
  geom_bar()
```

Now proceed to create a grouped bar chart by adding the sex variable too.
Try swapping the mappings between class and sex and observe how the 
information you are able to glean from the visualization changes.
Since class is an ordinal variable, use an appropriate color scale.

## Grouped Bar Chart

What we're really interested in is to see how survival is related to the
different categorical variables in the data set. We'll start simply by looking
at class and survival.

Construct a grouped bar chart displaying class and survival. Think carefully
about which variable you want to map to fill and swap the mappings to 
see the effect this has on the visualization.

```{r include = FALSE}
ggplot(titanic, aes(Class, fill = Survived)) +
  geom_bar(position = "dodge")
```

Can you see in which of the passenger classes a majority of the passengers
died?

Try to change the color manually for the the fill to some colors you think
are good fits for the data.

## Proportional Bar Charts

It is hard to compare survival among the groups due to the differences in sizes
of the groups. To get around this, plot a proportional stacked bar chart. 

```{r include = FALSE}
ggplot(titanic, aes(Class, fill = Survived)) +
  geom_bar(position = "fill")
```

Consider the advantages and downsides to using this type of plot.

## Grouped Proportional Bar Charts

During the presentation, we used the following code to prepare data
for plotting a grouped proportional bar chart.

```{r eval = FALSE}
happiness_props <- group_by(happiness, gender, happy) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n))
```

Adapt this code to the titanic data to plot a grouped, proportional bar
chart.

```{r include = FALSE}
titanic_props <- 
  group_by(titanic, Class, Survived) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n))

ggplot(titanic_props, aes(Class, prop, fill = Survived)) +
  geom_col(position = "dodge")
```

What do you think about this plot compared to the earlier ones? 

# Bonus: Mosaic Plots

If you have time over, visit the webpage for
[ggmosaic](https://haleyjeppson.github.io/ggmosaic/), which is an addon to
ggplot2 that lets you draw mosaic plots, which can be used as an alternative
to bar charts.

Then try to create a mosaic plot for the Titanic data for all of the variables
(age, sex, and survival status).
