---
title: "Practice: Multivariate Data"
subtitle: "Data Visualization with R and ggplot2"
author: "Johan Larsson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = FALSE,
  fig.width  = 5,
  fig.height = 4,
  dev        = "png"
)

options(scipen = 999)

library(tidyverse)
```

# Overview

In this quiz we will practice visualizing multiple variables using
aesthetics and facets.

For this session, you will need to install the following packages:

- [ggrepel](https://CRAN.R-project.org/package=ggrepel)
- [ggtern](https://CRAN.R-project.org/package=ggtern) (optional)

Do so (if you haven't already) by running the following code.

```{r eval = FALSE}
install.packages(c("ggrepel", "ggtern"))
```

# Star Wars

Let's return to the Star Wars data set (`starwars`) that we saw earlier!

Begin with a scatter plot of height versus mass. Map the sex of the characters
to color of the points and use a logarithmic scale for the y axis.
Change the color palette if you prefer.

```{r include = FALSE}
ggplot(starwars, aes(height, mass, color = sex)) +
  geom_point() +
  scale_y_log10()
```

One of the characters seem to stand out in terms of mass. To find out who this
outlier might be (which might not be hard to guess for anyone who is a Star Wars
fan) we'll map text to these observations. To do so, we'll use the **ggrepel**
package. Load the package.

```{r}
library(ggrepel)
```

Now add the `geom_text_repel()` geom to your previous plot. Map `name` to
`label`.

```{r include = FALSE}
ggplot(starwars, aes(height, mass, color = sex, label = name)) +
  geom_point() +
  scale_y_log10() +
  geom_text_repel()
```

*Note: we could have just used `geom_text()` or `geom_label()` instead here,
but the `geom_text_repel()` function from **ggrepel** automatically avoids
overlap in the plot, which is often very useful.*

# Adding Labels only to Some Observations

The plot nicely connects names to the points but is kind of messy. For data sets
where this happens, it might be worthwhile to consider whether it's possible to
reduce the amount of clutter by applying text labels only to some observations.

To do so, the best way is to create a separate, filtered data
set for the labels we want. Let's say that we want to label the observation in each level of `sex`
with the largest mass. To proceed, we group by
sex and slice the data set so that only the observations with the highest mass
in each group remain, then we `pull()` the `name` variable and store it in
`high_mass`.

```{r}
high_mass <-
  starwars %>%
  group_by(sex) %>%
  slice(which.max(mass)) %>%
  pull(name)
```

Next, we `mutate()` the data set by adding a new variable `highlight` that gives
the name of the observation if it corresponds to one of the observations in
`high_mass` and otherwise `""`, which results in no label.

```{r}
starwars2 <-
  starwars %>%
  mutate(highlight = ifelse(name %in% high_mass, name, ""))
```

Now we are ready to plot. Use the new `highlight` column as your `label`
mapping, but keep the plot the same otherwise.

```{r include = FALSE}
ggplot(starwars2, aes(height, mass, color = sex, label = highlight)) +
  geom_point() +
  scale_y_log10() +
  geom_text_repel()
```

Which character has the largest mass in the `NA` group?

- Sly Moore [correct]
- Grievous
- IG-88
- Beru Whitesun lars

# Faceting

Let's take a closer look at the relationship between height and mass across the
different star wars movies. The column `films` contains character vectors of
this, but to use it in a plot, we first need to `unnest()` this column.

Call the following line to unnest this column, creating a longer version of the
data set wherein each character may occur more than once.

```{r}
starwars3 <-
  unnest(starwars2, films)
```

Now create a new scatter plot, again with mass against height on the x and y
axes, but with facets corresponding to `gender` (note: not sex) and `films`.
Keep the scale transformation you used previously (logarithmic). You can also
keep the labels you used in your previous plot or, if you want a bonus task,
create new labels using the method above to highlight the heaviest character for
each film and gender.

Note that you can use both `facet_wrap()` and `facet_grid()` here. Try out both
and compare the results.

```{r include = FALSE}
ggplot(starwars3, aes(height, mass, color = sex, label = highlight)) +
  geom_point() +
  scale_y_log10() +
  geom_text_repel() +
  facet_grid(gender ~ films)
```

# Switzerland Demographics Revisited

Once more, we take a look at the data from provinces in Switzerland (stored in
the `swiss` data set), which you should be familiar with by now. Before we get
started, let's make a tibble out of the data.frame by calling the following
command.

```{r}
swiss2 <- as_tibble(swiss, rownames = "Province")
```

This time, create a bubble plot with `Agriculture` and `Examination` mapped to
the x and y axes respectively and `Fertility` mapped to the radii of the
bubbles. Also add text labels as before. Put the `size` mapping
inside the call to `geom_point()` to avoid mapping text size to fertility.

```{r include = FALSE}
ggplot(swiss2, aes(Agriculture, Examination, label = Province)) +
  geom_point(aes(size = Fertility), col = "steelblue4") +
  geom_text_repel()
```

What do you think of the bubbles in the plot? Try to guess what the values
are by looking at the size of the bubbles. As a bonus exercise, 
compute the right answers and see if you were correct.

# Mapping Color to Categories

Let's take this plot one step further. Let's create a new variable in the data
set indicating whether the province is mostly Catholic or mostly protestant and
then color the bubbles in your plot based on this variable. 

```{r}
swiss3 <-
  swiss2 %>%
  mutate(Religion = ifelse(Catholic >= 50, "Catholic", "Protestant"))

```

```{r include = FALSE}
ggplot(
  swiss3,
  aes(Agriculture, Examination, label = Province, color = Religion)
) +
  geom_point(aes(size = Fertility)) +
  geom_text_repel()
```

Which catholic province has the lowest proportion of workers working in
agriculture?

- Rive Gauche

## Bonus: Ternary Plots

Ternary plots can be used to visualize three variables in two-dimensional
display and are useful for compositional data when three features (variables)
sum to a constant, typically 1 (100%). They are common when studying geological
features, such as samples of rocks, when each sample is comprised of three
different types of minerals, for example.

To plot ternary plots with ggplot2, load the **ggtern** package, which you
should have installed by now.

```{r}
library(ggtern)
```

The package comes with a few data sets to showcase ternary plots and we'll use
the `Feldspar` data set here. Load it by calling `data(Feldspar)`. Now, to
construct a ternary plot, use the `ggtern()` function instead of `ggplot()`. It
is a wrapper to the latter, which provides a coordinate transformation that is
necessary to get reasonable results.

Then, to produce the plot, map `An` to the x axis, `Ab` to the y axis, and `Or`
to the z axis. Also map color to the feldspar variety (`Feldspar`).

```{r include = FALSE}
data(Feldspar)

ggtern(Feldspar, aes(An, Ab, z = Or, col = Feldspar)) +
    geom_point()
```

The resulting plot shows, for each sample of feldspar, its composition in terms
of the three constituents `Ab`, `Or`, and `An`, which represents chemical
compounds. To read the individual values for any point, you read the chart along
the diagonal or horizontal grid lines. Note also that the labels for the
components are positioned such that a observation close to it indicates that it
contains a lot of that component.

What is (roughly) the highest value of `Or` that any Feldspar in the group
"Plagioclase" takes? (A margin of +- 5 is tolerated).

- 20 (+- 5)
