---
title: "Practice: One Variable"
subtitle: "Data Visualization with R and ggplot2"
author: "Johan Larsson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
options(scipen = 999)
```

# Overview

In this session we will visualize the distribution of a single variable.

You need to install the following packages:

- [modeldata](https://CRAN.R-project.org/package=modeldata)
- [patchwork](https://CRAN.R-project.org/package=patchwork)

# Penguins 

Load **tidyverse** and the **modeldata** packages. Some packages directly
export their dataset but the **modeldata** package does not, which means me
need to call `data(penguins)` to make this package available to us. 

```{r}
library(tidyverse)
library(modeldata)

data(penguins)
```

Start by familiarizing yourself with the dataset. A good first step is to call
`head(penguins)` and `glimpse(penguins)`. Also read the documentation for the
dataset (although in this case it is not very revealing) by using the `F1`
shortcut in R Studio or by calling `?penguins`.

## Dotplot

Plot a dotplot of the `body_mass_g` variable in the penguins data using the
default settings by calling the following lines.

```{r fig.show = "hide"}
ggplot(penguins, aes(body_mass_g)) +
  geom_dotplot(binwidth = 100)
```

Now plot another dotplot using a binwidth of 1. What happens? Why is this setting
not appropriate for this data? Try some different values for `binwidth`. 

## Histogram

Continue by plotting the `body_mass_g` variable using a histogram with the default
settings. Recall that the layer function for a histogram is called
`geom_histogram()`. Use the same mappings as before.

```{r, include = FALSE}
ggplot(penguins, aes(body_mass_g)) +
  geom_histogram()
```

The default value is to use 30 bins for the histogram, which is not always
optimal. Try changing this by settings the `bins` and `binwidth` arguments in 
`geom_histogram()` to some other value. Observe how the histogram changes.

Next, we'll use a custom sequence of break points for our
histogram bins. Create your own vector of breakpoints using the
`c()` function and use this vector inside the call to
`geom_histogram()`. Supply this vector of break points to the `breaks`
argument.

```{r, include = FALSE}
breaks <- c(2500, 3500, 4000, 5000, 7200)

ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(breaks = breaks)
```

```{r, eval = FALSE}
# create a vector of break points
breaks <- c(NA)

# call ggplot, using your vector of break points
```

### Histogram Bin Rules-of-Thumb

There have been several attempts to produce rules of thumb for selecting
settings for either the number of bins or the bin width. Although you shouldn't
be using these blindly, they do take away a bit of the guesswork in selecting
bins and removes some of the possibility for bias. Go to
<https://en.wikipedia.org/wiki/Histogram#Number_of_bins_and_width> and read up
on the **Rice Rule**.

Compute the optimal number of bins for the Penguins data 
set according to the Rice Rule and try it out on the `body_mass_g` variable.

First, however, we compute the number of observations, taking care to drop
missing values. 

```{r}
n <- drop_na(penguins, body_mass_g) |>
  nrow()
```

Now use `n` to figure out what the optimal number of bins is according to this
rule, and then use this in your histogram. What do you think of the result?

*Hint: Note that $\sqrt[^3]{n} = n^\frac{1}{3}$.*

```{r, include = FALSE}
k <- 2 * n^(1/3)
```

```{r include = FALSE}
ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(bins = k)
```

## Box Plots

Now plot a box plot of the `body_mass_g` variable in the penguins data.

```{r, include = FALSE}
ggplot(penguins, aes(body_mass_g)) +
  geom_boxplot()
```

Try to figure out how many outliers there are in the plot, what the
median is, and the quartiles. What do you think of the boxplot compared
to the histogram?

## Density Plots

Let's try to use a density plot instead. First plot a density plot of body mass
in the penguins data using the default settings. Then try changing the
`bw` argument to something different (or use `adjust`, which acts as a
multiplier on the default value).

Now change the kernel for the density plot to a rectangular kernel,
using `kernel = "rectangular"` inside the call to `geom_density()`.

```{r, include = FALSE}
ggplot(penguins, aes(body_mass_g)) +
  geom_density(kernel = "rectangular")
```

What happens to the plot after you've switched the kernel to the rectangular
one? Do you understand why the result looks different?

# Patching Plots

The **patchwork** package is useful for combining separate ggplots into
one visualization. It's very easy to use yet is still flexible enough
to be able to generate a wide variety of layouts. First load the package.

```{r}
library(patchwork)
```

Now create three box plots, one for each of bill depth, bill length, and
flipper length for the penguins data set. Use the patchwork package to
combine the three box plots into one layout.

The easiest way to combine plots together with patchwork is to simply
add them together using `+`. Try to run the following code.

```{r eval = FALSE}
p <- ggplot(penguins)

p1 <- p + geom_boxplot(aes(y = bill_depth_mm))
p2 <- p + geom_boxplot(aes(y = bill_length_mm))
p3 <- p + geom_boxplot(aes(y = flipper_length_mm))

p1 + p2 + p3
```

Instead of `+`, you can use `|` too. In addition to these, you can
also use `/` and use parenthesis to group plots together. Try out some of
these on the plots above.

Let's say you want to stack `p1` and `p2` horizontally on top and
have `p3` on the bottom, spanning the entire width of the plot area. How
would you go about doing that?

If you have time over, go to [the patchwork
webpage](https://patchwork.data-imaginist.com/index.html) and check out
the *Guides* for more details about all the types of fun things you
can do with patchwork.
