---
title: "Practice: Two Variables and Scales"
subtitle: "Data Visualization with R and ggplot2"
author: "Johan Larsson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)

options(scipen = 999)

library(tidyverse)
```

# Overview

For this session you need the following packages:

- [tsibble](https://CRAN.R-project.org/package=tsibble)
- [MASS](https://CRAN.R-project.org/package=MASS)

If you haven't already installed them, use the following code to do so.

```{r, eval = FALSE}
install.packages(c("tsibble", "MASS"))
```

# Swiss Demographics


Let's take a look at the `swiss` data set. First, as always, browse the
documentation for the data set to get a sense of what the data concerns.

Let's start with a simple scatter plot of the proportion of Catholics
and fertility for each province.

```{r, eval = FALSE}
ggplot(swiss, aes(Catholic, Fertility)) +
  geom_point()
```

Try to explore some of the relationships in the data, focusing on
the relationship between fertility and the other variables in the data set.
For which relationships does the correlation appear to be strongest?
Try to describe the relationships you see. Are they linear, quadratic, or
something else?

# Miles per Gallon

Use the `mpg` data set to plot a scatter plot of number of cylinders (`cyl`)
against highway miles per gallon (`hwy`). 

```{r, include = FALSE}
ggplot(mpg, aes(cyl, hwy)) +
  geom_point()
```

There is a lot of overlap in the plot, which we should solve using jittering. 
Try adding jittering by specifying `position = position_jitter()` in
`geom_point()`. Set `height` and `width` in `position_jitter()` to some values.
Consider whether you want to jitter in both directions, or only in one of
them---which one in that case?

## Opacity 

Again plot a scatter plot of the number of cylinders against highway miles per
gallon using the `mpg` data set but this time use opacity (`alpha`) to
alleviate the effect of overlap.

```{r, include = FALSE}
ggplot(mpg, aes(cyl, hwy)) +
  geom_point(alpha = 0.2)
```

Try different combinations of `alpha` and jittering to settle on
a plot you find effective and easy to read.

## Box Plots

Make another plot of cylinders against highway miles per gallons, but this time
use a box plot instead. To do so, you need to first convert the cylinder
variable into a factor, and then construct the plot. (You can also do this
directly in the ggplot2 call, but then you need to modify the labels
accordingly.)

```{r}
mpg2 <- mutate(mpg, cyl = as_factor(cyl))
```

```{r include = FALSE}
mutate(mpg, cyl = as_factor(cyl)) %>%
  ggplot(aes(cyl, hwy)) +
    geom_boxplot()
```

What do you think about this plot compared to the scatter plot?

# Flow of River Nile

Run the following code to create a ggplot2-compatible data set out of the Nile
time series.

```{r}
library(tsibble)

nile <- as_tsibble(Nile) %>%
  rename(year = index, flow = value)
```

Read the documentation for the Nile data set and then construct
a plot to investigate the flow of the river across time.
Are there any seasonal trends in the plot?

# Mammals

Next, we'll look at some data on body and brain size of mammals; this time we'll
use the `mammals` data set from the **MASS** package. We make the data set
available by using the `data()` function in the following manner and then tidy
it up by turning the row names into its own column.

```{r}
data("mammals", package = "MASS")

# turn the rownames into a real variable
mammals_tidy <- rownames_to_column(mammals, "mammal")
```

*The reason that we are avoiding just calling `library(MASS)` here is that
**MASS** contains some functions that cause conflicts with functions from
**tidyverse**.*

Next, plot a standard scatter plot of body size against brain size using the
code below.

```{r mammals, eval = FALSE}
ggplot(mammals_tidy, aes(body, brain)) +
  geom_point()
```

As you can see, this plot does not really work too well. Try to fix this
problem by using different scales. Try to use `scale_x/y_log10()` and
`scale_x/y_sqrt()` (square root).

```{r eval = FALSE, echo = FALSE}
ggplot(mammals, aes(body, brain)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Are there any downsides to using non-linear scales?

## Direct Transformations

Instead of using scales, it's possible to use transform the data instead.
To, for instance, apply a log transformation to brain size and body size, we 
could do the following:

```{r mammals_direct, eval = FALSE}
ggplot(mammals_tidy, aes(log10(body), log10(brain))) +
  geom_point()
```

What is the difference when doing this compared to using the scale
functions? Which do you prefer (and why)?

If you have time left over, check out the documentation for
`scale_x/y_continuous()` at
<https://ggplot2.tidyverse.org/reference/scale_continuous.html> and try
out some other transformations by changing the value of the `trans` argument.
See the documentation for possible values. (There are a lot of them!)
