---
title: "Preparations"
author: "Johan Larsson"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    global_numbering: false
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 4,
  dev = "png"
)
```

Here are some instructions on how to setup your system properly for
the workshop, which involves installing R, R Studio, some R packages,
and making sure that everything is up-to-date.

## R {-}

Installing R is easy. Simply go to <https://cloud.r-project.org/> and download
the appropriate version of R depending on your operating system:

Linux
: Click on *Download R for Linux*, pick a link based on your distribution
  (such as `ubuntu/`), and then follow the instructions specific to that
  distribution.

Windows
: Click on *Download R for Windows*, then *base*, and finally
  *Download R X.X.X for Windows*.

Mac (OS) X
: Click on *Download R for (Mac) OS X*, then under "Latest release"
  click on *R-X.X.X.pkg*.

After this, you simply run the file you downloaded and follow the installation
guide.

## R Studio {-}

After you're done installing R, it's time to install R Studio. Simply go to
<https://rstudio.com/products/rstudio/download/#download> and pick a suitable
version for R Studio depending on your operating system. Download and install
it.

## Tidyverse {-}

During the workshop, we will be working mostly with packages that are
part of **tidyverse**: a collection of packages that work seamlessly
together. The package we will be using exclusively for our
visualizations, **ggplot2**, is part of the tidyverse.

To download and install all of the tidyverse packages, simply call

```{r, eval = FALSE}
install.packages("tidyverse")
```

in R Studio (in the terminal tab). 

It's possible that, when installing one or several of the packages,
you will get a message saying (approximately)

> There is a binary version available but the source version is later. Do
  you want to install from sources which need compilation?

In this case you can answer "no".

## Update Your Packages {-}

Finally, make sure that all your packages are up-to-date by
calling 

```{r, eval = FALSE}
update.packages()
```

and update everything.

Now you're good to go!
